\documentclass[%
  class = article,%
  crop = false,%
  float = true,%
  multi = true,%
  preview = false,%
]{standalone}
\onlyifstandalone{\usepackage[bookmarks,hyperindex=true,pdfencoding=auto,psdextra]{hyperref}}
\onlyifstandalone{\usepackage{ejb-dissertation}}
\begin{document}
I am surprised at how many people, places, and things are to blame for the creation of this dissertation. The first is my advisor, Daniel Secretary Lambrecht, who never got frustrated with me when I wasn't being productive, and mostly left me to my own devices, presumably because he didn't want to deal with me not knowing Vim and gnuplot. I will miss our discussions about how woodwind instruments are different from brass, how perfect the San Francisco weather is, how confusing the \texttt{NBasis/NBas6D/NB2/NB2car} mess is in \textsc{Q-Chem}, and what directions quantum chemistry is and should be heading in. I promise to never use B3LYP/6-31G** for anything ever again. More importantly, thank you for countering my occasionally poor communication skills and negative outlook with positive words of encouragement. I also want to thank Sean Garrett-Roe, who aside from always trying to make a collaboration with experimentalists as painless as possible, perhaps surpasses even me with fervent disgust toward Microsoft Word and was always a willing ear for my rants about computational chemists not using \LaTeX{}. You are a model for how well-rounded a scientist should be that I strove to emulate. I would like to thank Ken Jordan for supervising me during my first summer at Pitt, and teaching me that some printers may prefer to work in atomic units.

I am indebted to Thomas Brinzer for all the time and effort he put into our collaboration, trying to get me to understand multidimensional spectroscopy, and his always thoughtful critique of my work. You are the model for how a graduate student should be that I strove to emulate.

I would like to thank Evgeny Epifanovsky and \textsc{Q-Chem} for giving me a two month \st{vacation}internship in Pleasanton, which was the most valuable and enjoyable experience of my PhD.

I would like to thank Daniel Smith for inviting me to a MolSSI workshop and for baiting me with authorship on the \textsc{Psi4NumPy} paper after watching me hack on \textsc{Psi4} in a Twitch stream.

As a strong believer in learning one's tools, I would like to thank Emacs for making me faster at editing text than Ben Albrecht, who learned with me that people shouldn't share flash drives when printing posters. I would like to thank Kevin Gasperich, Shiv Upadhyay, Amanda Dumi, and Daniel Burrill for joining me at weekly programming study groups that often got derailed as they learned how to push my buttons. I would like to thank Keith Werling for being my original partner-in-crime as Daniel's other first student (though you are really the zeroth). I would like to thank Adam Gagorik for teaching me how to use Matplotlib and setting and example for getting work done as a graduate student. I would like to thank Daniel Bolt, Tyler Rohrs, and Halina Werner for tolerating some truly bad bowling strings while my mind was in other places, talking coffee, going to the symphony, and attending my own performances. I also acknowledge Forward Lanes, Row House Cinema, Hollywood Theater, Voodoo Brewery, and Commonplace Coffee for providing places to go other than work.

I acknowledge \href{https://chemistry.stackexchange.com/users/194/}{\color{black}{Chemistry Stack Exchange}} for being incredibly distracting but also providing material and Easter eggs that made it into this dissertation. In particular, I would like to thank Brian Skinn and Zhe Lu for our lunch meetings and all the encouragement that I could make it out of academia. I would also like to thank cclib and particularly Karol Langer for allowing me to contribute to the project and helping me enter the world of open-source software development.

There are people from home who helped me as well. I would like to thank Car Chat for coming to visit me and welcoming me every time I came home, especially Brendan Reilly, who introduced me to anime and indirectly got me started playing music again. I want to thank Garen Chiloyan for listening to me complain about almost everything since the start of undergrad, sharing grad school gossip, and agreeing that Massachusetts really is the best. I would like to thank Marshall Brennan for explaining what I should expect in grad school and being right about 99\% of it.

I want to thank Adam Pratt and Tuguldur Togo Odbadrakh (Togie Stogie? Tuguldur Schmuguldur? Togo Schmogo?) for our roller coaster ride, and I apologize for showing you my actual bad side.

My degree would not have been completed if not for the massive amount of music I listened to, primarily from Alfred Brendel, Francesco Corti, Passepied, and Judy Bedford, my original bassoon instructor who still inspires me. One of my biggest mistakes was not playing music for my first five years here, so I would like to thank Brad Townsend, Barbara Hois, and Roger Zahab for taking a chance and allowing me to play in ensembles. I especially want to thank Linda Fisher for her patience while teaching me how to play and make reeds again.

Finally but most importantly, I am eternally thankful to my parents, Gavra and John. You are my most passionate supporters, loudest critics, and best friends. Only you understand how difficult this was for me. I love you.
\end{document}
